# STATISTICAL ANALYSIS AND PLOTS

```{r}
#Check for required packages and install them if they are not already available
list.of.packages <- c("readr", "data.table","tidyverse","flextable","dplyr",
                      "scales","stats","rstatix","dslabs","ggplot2",
                      "geomtextpath","ggrain","ggh4x","ggpubr","ggthemes","plotROC",
                      "pROC","verification", "PRROC"
                      )
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)

if (!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes")
}
remotes::install_github("thomas-neitmann/mdthemes", upgrade = "never")

library(readr)
library(data.table)
library(tidyverse)
library(flextable)
library(dplyr)
library(scales)
library(mdthemes) #Installed directly from github
library(stats)
library(rstatix)
library(dslabs)
library(ggplot2)
library(geomtextpath) #Text in the graphs of ggplot (used in geom_denstiy)
library(ggrain) #Axis manipulation
library(ggh4x) #Manual axis scales in facet wrap (ggplot)
library(ggpubr) #Add p-values manualy in ggplot
library(ggthemes)
library(plotROC) #ROC curves (with ggplot)
library(pROC) #ROC curve analysis 
library(verification) #Statistics of the ROC curves
library(PRROC) #Precision-Recall ROC curves

#Set working directory
setwd("/path/to/my/github/directory")
set.seed(42) #Seed for reproducibility
```

## BACTERIAL COUNT ANALYSIS

### Figure 1. Bacterial count test

Bacterial count (bacteria/µL) of stool samples: analysis

```{r}
#Flow cytometry tests --> Bacterial count (bacteria/µL)
bct_count_test <- read_csv("Data/Bacterial_count_tests.csv", col_names = TRUE, col_types = NULL)

#R-squared values for each function (by sample)
r_squared <- numeric()

#Fit regression model for each sample
for (sample_name in unique(bct_count_test$Sample_name)) {
  sample_data <- filter(bct_count_test, Sample_name == sample_name)
  model <- lm(Bacterial_count ~ Dilution, data = sample_data)
  r_squared <- c(r_squared, summary(model)$r.squared)
}

#Mean of the R-squared values 
mean_r2 <- mean(r_squared) #Result = 0.997918.
```

Code for the plot of figure 1:

```{r}
#We create a function to convert the default scientific notation to a format that we like. Then we will incorporate this function in 'scale x continuous'
scientific <- function(x){
  ifelse(x==0, "0", parse(text = gsub("[+]", "", gsub("e", "%*% 10^", scientific_format()(x)))))
}

#Plot FIGURE 2
figure_1_plot <- ggplot(bct_count_test, aes(x = Dilution, y = Bacterial_count, colour = Sample_name)) +
  geom_point(size = 0.7, show.legend = FALSE, alpha = 0.5) +
  geom_smooth(method = "lm", formula = 'y ~ x', span = 0.8, linewidth = 0.4, alpha = .15, aes(fill = Sample_name)) +
  scale_color_brewer(palette = "Greys") +
  scale_fill_brewer(palette = "Greys") +
  scale_y_continuous(limits = c(0, 3500)) + 
  scale_x_continuous(breaks = c(0, 0.0001, 0.0002, 0.001), label = scientific) +
  theme_minimal() +
  labs(x = "Stool sample dilution", y = "Bacterial count (bacteria/µL)") +
  annotate("text", x = 0.0004, y = 2750, label = "italic(R) ^ 2 == 0.9979", parse = TRUE, color = "gray40", size = 10) +
  theme(legend.position = "none", 
        axis.text = element_text(size = 25),  # Size of axis labels (numbers)
        axis.title.y = element_text(vjust = +3, size = 30), #Size of axis titles 
        axis.title.x = element_text(vjust = -1, size = 30), 
        strip.text.x = element_blank(), 
        plot.margin = margin(0.2, 0.2, 0.2, 0.2, "in")) #Margins of the plot 

figure_1_plot

#ggsave("Figures/Figure_1_bct_test.jpeg", width = 16, height = 9, units = "in", dpi = 300)
```

### Figure 2. Density plot bacterial count

Density plot of the bacterial counts (bacteria/µL) of the original sample.

```{r}
#Bacterial count results of the original sample (bacteria/µL) --> Healthy donors and dysbiotic patients with CDI
total_bacterial_count <- read_csv("Data/Bacterial_count_total.csv", col_names = TRUE, col_types = NULL)

#Scientific notation for the x axis of the plot --> 'scale x continuous'
scientific <- function(x){
  ifelse(x==0, "0", parse(text = gsub("[+]", "", gsub("e", "%*% 10^", scientific_format()(x)))))
}

#Plot FIGURE 2
figure_2_plot_outlier <- ggplot(total_bacterial_count, aes(x = bacterial_count, fill = patient_type, label = label_sample)) + 
  geom_density(alpha = 0.5, color = "dimgray", linewidth = 0) + 
  geom_textdensity(linewidth = 0.5, color = "gray30", size = 4, hjust = 0.36, spacing = 40) +
  labs(x = "Bacterial count (bacteria/µL)", y = "Density", title = "Bacterial count with the outlier in CDI patients") +
  scale_x_continuous(label = scientific) + 
  scale_y_continuous(label = scientific) +
  scale_fill_manual(values = c("indianred3", "slategray3")) +
  labs(x = "Bacterial count (bacteria/µL)", y = NULL, fill = NULL) + 
  theme_minimal() + 
  theme(legend.position = "none", 
        axis.text = element_text(size = 10),  # Size of axis labels
        axis.title.y = element_text(vjust = +3, size = 10), #Size of axis titles 
        axis.title.x = element_text(vjust = -1, size = 10), 
        strip.text.x = element_blank(), 
        plot.margin = margin(0.2, 0.2, 0.2, 0.2, "in")) #Margins of the plot 

figure_2_plot_outlier 
```

As the plot shows an outlier in the bacterial count of dysbiotic patients with CDI, we deleted the sample and did not consider it in posterior analysis (of both bacterial count and SCFAs concentrations):

```{r}
bacterial_count <- total_bacterial_count[-138,]

figure_2_plot <- ggplot(bacterial_count, aes(x = bacterial_count, fill = patient_type, label = label_sample)) + 
  geom_density(alpha = 0.5, color = "dimgray", linewidth = 0) + 
  geom_textdensity(linewidth = 1, color = "gray30", size = 8, hjust = 0.36, spacing = 40) +
  labs(x = "Bacterial count (bacteria/µL)", y = "Density") +
  scale_x_continuous(label = scientific) + 
  scale_y_continuous(label = scientific) +
  scale_fill_manual(values = c("indianred3", "slategray3")) +
  theme_minimal() + 
  theme(legend.position = "none", 
        axis.text = element_text(size = 25),  # Size of axis labels
        axis.title.y = element_text(vjust = +3, size = 30), #Size of axis titles 
        axis.title.x = element_text(vjust = -1, size = 30), 
        strip.text.x = element_blank(), 
        plot.margin = margin(0.2, 0.2, 0.2, 0.2, "in")) #Margins of the plot 

figure_2_plot

#ggsave("Figures/Figure_2_bct_density.jpeg", width = 16, height = 9, units = "in", dpi = 300)
```

### Shapiro-Wilk Normality Test

Statistical analysis (normality tests):

```{r}
#Test Shapiro-Wilk for normality
shapiro.test(subset(bacterial_count, patient_type == "healthy_donor")$bacterial_count)
#Normal distribution --> p-value = 0.1381

shapiro.test(subset(bacterial_count, patient_type == "cdiff_patient")$bacterial_count)
#Not a normal distribution --> p-value = 4.328e-05
```

### Mann Whitney U Test

Comparison of the bacterial count results by type of patient:

```{r}
#Comparison between C.diff and HD samples (bacterial count)
stat.test_bct_count <- bacterial_count %>% wilcox_test(bacterial_count ~ patient_type)

#Signifficant differences --> p-value = 5.57e-19
```

## SAMPLE STABILITY OF SAMPLES FROM PATIENTS WITH CDI

### Friedman test

*C.difficile* sample stability in the fridge (4ºC). Comparison of the bacterial count stability:

```{r}
stability_cdiff <- read_csv("Data/Cdiff_sample_stability.csv", col_names = TRUE, col_types = NULL) 

#Bacterial count average for stability tests 
bacterial_count_stability <- dplyr::select(stability_cdiff, sample_name, week, Bacterial_count_1, Bacterial_count_2, Bacterial_count_3) %>% mutate(average_bct_count = rowMeans(.[, c("Bacterial_count_1", "Bacterial_count_2", "Bacterial_count_3")])) %>% dplyr::select(sample_name, week, average_bct_count) 

#Friedman test -> Comparison of the bacterial count between weeks 
stat.test_bct_stability <- bacterial_count_stability %>% friedman_test(average_bct_count ~ week | sample_name)

#p-value of 0.45; no significant differences between groups. 
```

Comparison of the SCFA concentration stability:

```{r}
#SCFA concentrations  
scfa_stability <- dplyr::select(stability_cdiff, sample_name, week, Acetate_uM, Propionate_uM, Butyrate_uM)

#Friedman test -> Comparison of the acetate, propionate and butyrate concentrations
stat.test_acetate_stability <- scfa_stability %>% friedman_test(Acetate_uM ~ week | sample_name)
stat.test_propionate_stability <- scfa_stability %>% friedman_test(Propionate_uM ~ week | sample_name)
stat.test_butyrate_stability <- scfa_stability %>% friedman_test(Butyrate_uM ~ week | sample_name)

#Generate a summary table
table_friedman <- rbind(stat.test_acetate_stability,stat.test_propionate_stability,stat.test_butyrate_stability)
colnames(table_friedman[1]) <- "SCFA"
table_friedman

#p-value of 0.615 (acetate), 0.042 (propionate), 0.457 (butyrate); no significant differences between groups (alfa = 0.01).
```

## NORMALISATION OF SCFAs

Data load from healthy donors (healthy_donor) and dysbiotic patients with CDI infections (cdiff_patient):

```{r}
#Data load 
#SCFAs levels (µg/mL) HEALTHY DONORS
scfa_levels_hd <- read_csv("Data/SCFA_levels_ugmL_HD.csv") 
#SCFAs levels (µg/mL) DYSBIOTIC PATIENTS
scfa_levels_cdiff <- read_csv("Data/SCFA_levels_ugmL_CDIFF.csv") 
```

SCFAs levels (µg/mL) of stool samples of healthy donors (healthy_donor):

```{r}
#HEALTHY DONORS
#Normalisation factor calculation
norm_factors_hd <- bacterial_count %>%
  filter(patient_type == "healthy_donor") %>%
  mutate(norm_fctr_hd = bacterial_count / max(bacterial_count))

#Levels without normalisation
scfa_notnorm_hd <- scfa_levels_hd %>%
  mutate(normalisation = "not_normalised", scfa_concentration = scfa_ugmL) %>%
  dplyr::select(sample_number, patient_type, scfa, normalisation, scfa_concentration)

#Normalisation by the bacterial count
scfa_bct_hd <- scfa_levels_hd %>%
  mutate(normalisation = "normalised_bct", 
         scfa_concentration = scfa_ugmL * norm_factors_hd$norm_fctr_hd) %>%
  dplyr::select(sample_number, patient_type, scfa, normalisation, scfa_concentration)

#Normalisation by fresh weight
scfa_weight_hd <- scfa_levels_hd %>%
  mutate(normalisation = "normalised_weight", 
         scfa_concentration = scfa_ugmL / fresh_weight) %>%
  dplyr::select(sample_number, patient_type, scfa, normalisation, scfa_concentration)
```

SCFAs levels (µg/mL) of stool samples of dysbiotic patients with CDI (cdiff_patients):

```{r}
#DYSBIOTIC PATIENTS (C. difficile infection)
#Normalisation factor calculation
norm_factors_cdiff <- bacterial_count %>%
  filter(patient_type == "cdiff_patient") %>%
  mutate(norm_fctr_cdiff = bacterial_count / max(bacterial_count))

#Levels without normalisation
scfa_notnorm_cdiff <- scfa_levels_cdiff %>%
  mutate(normalisation = "not_normalised", scfa_concentration = scfa_ugmL) %>%
  dplyr::select(sample_number, patient_type, scfa, normalisation, scfa_concentration)

#Normalisation by the bacterial count (scfa_bct)
scfa_bct_cdiff <- scfa_levels_cdiff %>%
  mutate(normalisation = "normalised_bct", 
         scfa_concentration = scfa_ugmL * norm_factors_cdiff$norm_fctr_cdiff) %>%
  dplyr::select(sample_number, patient_type, scfa, normalisation, scfa_concentration)

#Normalisation by fresh weight (scfa_weight)
scfa_weight_cdiff <- scfa_levels_cdiff %>%
  mutate(normalisation = "normalised_weight", 
         scfa_concentration = scfa_ugmL / fresh_weight) %>%
  dplyr::select(sample_number, patient_type, scfa, normalisation, scfa_concentration)
```

Combination of the data frames:

```{r}
#Combination of the data frames from healthy donors and dysbiotic patients
scfa_total_levels <- bind_rows(scfa_notnorm_hd, scfa_bct_hd, scfa_weight_hd, scfa_notnorm_cdiff, scfa_bct_cdiff, scfa_weight_cdiff)
```

Data summary:

```{r}
#Summarize data (grouping then summarizing in data.table)
scfa_total_levels <- setDT(scfa_total_levels)
total_summary <- scfa_total_levels[, .(median = median(scfa_concentration), IQR = IQR(scfa_concentration), Q1 = quantile(scfa_concentration, 0.25), Q3 = quantile(scfa_concentration, 0.75), min = min(scfa_concentration), max = max(scfa_concentration)), by = "normalisation,scfa,patient_type"]

total_summary
```

### Figure 3. Boxplots

Code for the plot of figure 3:

#### A) Acetate

```{r}
#ACETATE
acetate_data <- scfa_total_levels[scfa_total_levels$scfa=="acetate",]

#Reorder the levels of the 'patient_type' factor variable (so we have the healthy donors first in the boxplot)
acetate_data$patient_type <- factor(acetate_data$patient_type, levels = c("healthy_donor", "cdiff_patient"))

#Comparison between C.diff and HD samples
stat.test_acetate <- acetate_data %>% group_by(normalisation, scfa) %>% wilcox_test(scfa_concentration ~ patient_type) %>% add_significance()

#We define the place of the significance bars in the test we previously performed
#Define the y postition (note that the order is bct > weight > not norm)
stat.test_acetate <- stat.test_acetate %>% add_xy_position(x = "scfa", dodge = 0.8) %>% mutate(y.position = c(730, 7.8, 785))

#Define scales for each facet level
scales_y_axis_acetate <- list(
  "not_normalised" = scale_y_continuous(limits = c(0, 790), breaks = c(0, 250, 500, 750)),
  "normalised_weight" = scale_y_continuous(limits = c(0, 7.9), breaks = c(0, 2.5, 5, 7.5)),
  "normalised_bct" = scale_y_continuous(limits = c(0, 790), breaks = c(0, 250, 500, 750)))

#BOXPLOT ACETATE
figure_3A_plot <- ggplot(acetate_data, aes(x = scfa, y = scfa_concentration, colour = patient_type)) +
  facet_wrap(~factor(normalisation, levels = c("not_normalised", "normalised_weight", "normalised_bct")), labeller = as_labeller(c("not_normalised" = "Not normalised\n(µg/mL)", "normalised_bct" = "Normalised by bacterial count\n(normalised concentration)", "normalised_weight" = "Normalised by fresh weight\n(µg/mg)")), scales = "free_y") +
  facetted_pos_scales(y = scales_y_axis_acetate) +
  scale_colour_manual(values = c("healthy_donor" = "skyblue4", "cdiff_patient" = "indianred4"),
                      name = NULL,
                      labels = c("Healthy donors (n = 115)", "Dysbiotic patients (n = 40)")) + 
  stat_pvalue_manual(stat.test_acetate, label = "p.signif", tip.length = 0.01, hide.ns = FALSE, size = 6) + 
  geom_boxplot(width = 0.5, alpha = 0.5, lwd = 1, outliers = FALSE) + #To avoid overlay with the points
  geom_point(position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.5, seed = 42), aes(group = patient_type), alpha = 0.5, size = 2) +
  labs(x = NULL, y = "Acetate levels", tag = "A)") +
  scale_x_discrete(label = NULL) +
  theme_light() +
  theme(text = element_text(size = 25),
        axis.text = element_text(size = 18),  # Size of axis labels (numbers)
        axis.title.y = element_text(vjust = +3, size = 30), #Size of axis titles 
        plot.margin = margin(0.2, 2, 0.2, 0.2, "in"), #Margins of the plot (top, right, bottom, left) 
        legend.position = "none", 
        axis.ticks.x=element_blank(),
        legend.text=element_text(size=rel(1)), 
        strip.text = element_text(color = "gray20"))
  
figure_3A_plot

#ggsave("Figures/Figure_3A_boxplot_acetate.jpeg", width = 16, height = 9, units = "in", dpi = 300)
```

#### B) Propionate

```{r}
#PROPIONATE
propionate_data <- scfa_total_levels[scfa_total_levels$scfa=="propionate",]

#Reorder the levels of the 'patient_type' factor variable (so we have the healthy donors first in the boxplot)
propionate_data$patient_type <- factor(propionate_data$patient_type, levels = c("healthy_donor", "cdiff_patient"))

#Comparison between C.diff and HD samples
stat.test_propionate <- propionate_data %>% group_by(normalisation, scfa) %>% wilcox_test(scfa_concentration ~ patient_type) %>% add_significance()

#We define the place of the significance bars in the test we previously performed
#Define the y postition (note that the order is bct > weight > not norm)
stat.test_propionate <- stat.test_propionate %>% add_xy_position(x = "scfa", dodge = 0.8) %>% mutate(y.position = c(260, 3.3, 330))

#Define scales for each facet level
scales_y_axis_propionate <- list(
  "not_normalised" = scale_y_continuous(limits = c(0, 490), breaks = c(0, 150, 300, 450)),
  "normalised_weight" = scale_y_continuous(limits = c(0, 4.9), breaks = c(0, 1.5, 3, 4.5)),
  "normalised_bct" = scale_y_continuous(limits = c(0, 490), breaks = c(0, 150, 300, 450)))

#BOXPLOT PROPIONATE
figure_3B_plot <- ggplot(propionate_data, aes(x = scfa, y = scfa_concentration, colour = patient_type)) +
  facet_wrap(~factor(normalisation, levels = c("not_normalised", "normalised_weight", "normalised_bct")), labeller = as_labeller(c("not_normalised" = "Not normalised\n(µg/mL)", "normalised_bct" = "Normalised by bacterial count\n(normalised concentration)", "normalised_weight" = "Normalised by fresh weight\n(µg/mg)")), scales = "free_y") +
  facetted_pos_scales(y = scales_y_axis_propionate) +
  scale_colour_manual(values = c("healthy_donor" = "skyblue4", "cdiff_patient" = "indianred4"),
                      name = NULL,
                      labels = c("Healthy donors (n = 115)", "Dysbiotic patients (n = 40)")) + 
  stat_pvalue_manual(stat.test_propionate, label = "p.signif", tip.length = 0.01, hide.ns = FALSE, size = 6) + 
  geom_boxplot(width = 0.5, alpha = 0.5, lwd = 1, outliers = FALSE) + #To avoid overlay with the points
  geom_point(position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.5, seed = 42), aes(group = patient_type), alpha = 0.5, size = 2) +
  labs(x = NULL, y = "Propionate levels", tag = "B)") +
  scale_x_discrete(label = NULL) +
  theme_light() +
  theme(text = element_text(size = 25),
        axis.text = element_text(size = 18),  # Size of axis labels (numbers)
        axis.title.y = element_text(vjust = +3, size = 30), #Size of axis titles 
        plot.margin = margin(0.2, 2, 0.2, 0.2, "in"), #Margins of the plot (top, right, bottom, left) 
        legend.position = "none", 
        axis.ticks.x=element_blank(),
        legend.text=element_text(size=rel(1)), 
        strip.text = element_text(color = "gray20"))
  
figure_3B_plot

#ggsave("Figures/Figure_3B_boxplot_propionate.jpeg", width = 16, height = 9, units = "in", dpi = 300)
```

#### C) Butyrate

```{r}
#BUTYRATE
butyrate_data <- scfa_total_levels[scfa_total_levels$scfa=="butyrate",]

#Reorder the levels of the 'patient_type' factor variable (so we have the healthy donors first in the boxplot)
butyrate_data$patient_type <- factor(butyrate_data$patient_type, levels = c("healthy_donor", "cdiff_patient"))

#Comparison between C.diff and HD samples
stat.test_butyrate <- butyrate_data %>% group_by(normalisation, scfa) %>% wilcox_test(scfa_concentration ~ patient_type) %>% add_significance()

#We define the place of the significance bars in the test we previously performed
#Define the y postition (note that the order is bct > weight > not norm)
stat.test_butyrate <- stat.test_butyrate %>% add_xy_position(x = "scfa", dodge = 0.8) %>% mutate(y.position = c(280, 4.2, 420))

#Define scales for each facet level
scales_y_axis_butyrate <- list(
  "not_normalised" = scale_y_continuous(limits = c(0, 490), breaks = c(0, 150, 300, 450)),
  "normalised_weight" = scale_y_continuous(limits = c(0, 4.9), breaks = c(0, 1.5, 3, 4.5)), 
  "normalised_bct" = scale_y_continuous(limits = c(0, 490), breaks = c(0, 150, 300, 450)))

#BOXPLOT BUTYRATE
figure_3C_plot <- ggplot(butyrate_data, aes(x = scfa, y = scfa_concentration, colour = patient_type)) +
  facet_wrap(~factor(normalisation, levels = c("not_normalised", "normalised_weight", "normalised_bct")), labeller = as_labeller(c("not_normalised" = "Not normalised\n(µg/mL)", "normalised_bct" = "Normalised by bacterial count\n(normalised concentration)", "normalised_weight" = "Normalised by fresh weight\n(µg/mg)")), scales = "free_y") +
  facetted_pos_scales(y = scales_y_axis_butyrate) +
  scale_colour_manual(values = c("healthy_donor" = "skyblue4", "cdiff_patient" = "indianred4"),
                      name = NULL, 
                      labels = c("Healthy donors (n = 115)", "Dysbiotic patients (n = 40)")) + 
  stat_pvalue_manual(stat.test_butyrate, label = "p.signif", tip.length = 0.01, hide.ns = FALSE, size = 6) + 
  geom_boxplot(width = 0.5, alpha = 0.5, lwd = 1, outliers = FALSE) + #To avoid overlay with the points
  geom_point(position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.5, seed = 42), aes(group = patient_type), alpha = 0.5, size = 2) +
  labs(x = NULL, y = "Butyrate levels", tag = "C)") +
  scale_x_discrete(label = NULL) +
  theme_light() +
  theme(text = element_text(size = 25),
        axis.text = element_text(size = 18),  # Size of axis labels (numbers)
        axis.title.y = element_text(vjust = +3, size = 30), #Size of axis titles 
        plot.margin = margin(0.2, 2, 0.2, 0.2, "in"), #Margins of the plot (top, right, bottom, left) 
        legend.position = "bottom", 
        axis.ticks.x=element_blank(),
        legend.text=element_text(size=rel(1)), 
        strip.text = element_text(color = "gray20"))
  
figure_3C_plot

#ggsave("Figures/Figure_3C_boxplot_butyrate.jpeg", width = 16, height = 9, units = "in", dpi = 300)
```

## DENSITY PLOTS AND ROC CURVES

### ROC curve statistical analysis

Filtrate the SCFAs levels normalised by the bacterial count, and create the subset of data for each SCFA:

```{r}
#We filtrate data normalised by the bacterial count and add a column of the ROC outcome 
#Assuming --> 1 = Healthy donors; 0 = Ill patients 
scfa_bct_normalised <- scfa_total_levels %>% filter(normalisation == "normalised_bct") %>% mutate(roc_outcome = rep(c(1, 0), times = c(345, 120)))

#ROC curve analysis
#We filtrate the data of each SCFA and add the labels for the following density plots 
acetate_data_bct <- scfa_bct_normalised[scfa_bct_normalised$scfa=="acetate",] %>% mutate(labels_density = rep(c("Healthy donors", "Dysbiotic patients"), times = c(115, 40)))

propionate_data_bct <- scfa_bct_normalised[scfa_bct_normalised$scfa=="propionate",] %>% mutate(labels_density = rep(c("Healthy donors", "Dysbiotic patients"), times = c(115, 40)))

butyrate_data_bct <- scfa_bct_normalised[scfa_bct_normalised$scfa=="butyrate",] %>% mutate(labels_density = rep(c("Healthy donors", "Dysbiotic patients"), times = c(115, 40)))
```

Statistical analysis of the ROC curve for each SCFAs:

```{r}
#ROC parameters --> ACETATE 
roc_obj_acetate <- roc(response = acetate_data_bct$roc_outcome, 
                       predictor = acetate_data_bct$scfa_concentration)

#pvalue
roc_area_acetate <- roc.area(as.numeric(as.vector(acetate_data_bct$roc_outcome)), acetate_data_bct$scfa_concentration)
roc_area_acetate$p.value #p-value = 0.0001136763

roc_parameters_acetate <- coords(roc_obj_acetate, "best", ret =c("threshold", "specificity", "sensitivity", "precision", "youden", "closest.topleft"), transpose = FALSE) %>% mutate(auc = roc_obj_acetate$auc, ci_lower_bound = ci(roc_obj_acetate, of = "auc", method = "delong")[1], ci_upper_bound = ci(roc_obj_acetate, of = "auc", method = "delong")[3]) %>% mutate(pvalue = roc_area_acetate$p.value)

#ROC parameters --> PROPIONATE 
roc_obj_propionate <- roc(response = propionate_data_bct$roc_outcome, 
                       predictor = propionate_data_bct$scfa_concentration)

#pvalue
roc_area_propionate <- roc.area(as.numeric(as.vector(propionate_data_bct$roc_outcome)), propionate_data_bct$scfa_concentration)
roc_area_propionate$p.value #p-value = 7.819531e-07

roc_parameters_propionate <- coords(roc_obj_propionate, "best", ret =c("threshold", "specificity", "sensitivity", "precision", "youden", "closest.topleft"), transpose = FALSE) %>% mutate(auc = roc_obj_propionate$auc, ci_lower_bound = ci(roc_obj_propionate, of = "auc", method = "delong")[1], ci_upper_bound = ci(roc_obj_propionate, of = "auc", method = "delong")[3]) %>% mutate(pvalue = roc_area_propionate$p.value)

#ROC parameters --> BUTYRATE 
roc_obj_butyrate <- roc(response = butyrate_data_bct$roc_outcome, 
                       predictor = butyrate_data_bct$scfa_concentration)

#pvalue
roc_area_butyrate <- roc.area(as.numeric(as.vector(butyrate_data_bct$roc_outcome)), butyrate_data_bct$scfa_concentration)
roc_area_butyrate$p.value #p-value = 6.820517e-12

roc_parameters_butyrate <- coords(roc_obj_butyrate, "best", ret =c("threshold", "specificity", "sensitivity", "precision", "youden", "closest.topleft"), transpose = FALSE) %>% mutate(auc = roc_obj_butyrate$auc, ci_lower_bound = ci(roc_obj_butyrate, of = "auc", method = "delong")[1], ci_upper_bound = ci(roc_obj_butyrate, of = "auc", method = "delong")[3]) %>% mutate(pvalue = roc_area_butyrate$p.value)
```

Comparison between the AUC of the ROC curves:

```{r}
#Comparison between AUC of the three biomarkers (acetate, propionate, and butyrate). Provides valuable insights into their discriminatory power or predictive performance for a given outcome

#Perform DeLong test
test_result_auc_AP <- roc.test(roc_obj_acetate, roc_obj_propionate, method = "delong", reuse.auc = FALSE)
print(test_result_auc_AP) #p-value = 0.01432 (difference in AUC, if alfa = 0.05)

test_result_auc_AB <- roc.test(roc_obj_acetate, roc_obj_butyrate, method = "delong", reuse.auc = FALSE)
print(test_result_auc_AB) #p-value = 5.371e-07 (difference in AUC)

test_result_auc_PB <- roc.test(roc_obj_propionate, roc_obj_butyrate, method = "delong", reuse.auc = FALSE)
print(test_result_auc_PB) #p-value = 0.0002964 (difference in AUC)
```

### Figure 4.

Code for the plot of figure 4:

#### A) Acetate

```{r}
#ACETATE
#Check if follows a normal distribution 
#Healthy donors 
shapiro.test(subset(acetate_data_bct, patient_type == "healthy_donor")$scfa_concentration) 
#NOT a normal distribution (p-value = 9.375e-10)

#Dysbiotic patients
shapiro.test(subset(acetate_data_bct, patient_type == "cdiff_patient")$scfa_concentration) 
#NOT a normal distribution (p-value = 3.007e-06)

#DENSITY PLOT ACETATE
figure_4A_plot <- ggplot(acetate_data_bct, aes(x = scfa_concentration, fill = patient_type, label = labels_density)) + 
  geom_density(alpha = 0.3, color = "dimgray", linewidth = 0)  +
  geom_vline(xintercept = roc_parameters_acetate$threshold, color = "dimgray", linetype = "dashed", linewidth = 1) +
  geom_textdensity(linewidth = 1, color = "gray30", size = 8, hjust = 0.31, spacing = 40) +
  scale_fill_manual(values = c("healthy_donor" = "#443A83FF", "cdiff_patient" = "#440154FF")) +
  labs(x = "Acetate levels normalised by bacterial count", y = "Density", fill = NULL, tag = "A)") +
  scale_x_continuous(limits = c(0, 750), breaks = c(0, 150, 300, 450, 600, 750)) +
  scale_y_continuous(limits = c(0, 0.006), breaks = c(0, 0.0015, 0.003, 0.0045, 0.006)) +
  theme_minimal() +
  theme(legend.position = "none",
        text = element_text(size = 25),
        axis.text = element_text(size = 25),  # Size of axis labels
        axis.title.y = element_text(vjust = +3, size = 30), #Size of axis titles 
        axis.title.x = element_text(vjust = -1, size = 30), 
        strip.text.x = element_blank(), 
        plot.margin = margin(0.2, 0.2, 0.2, 0.2, "in")) #Margins of the plot 

figure_4A_plot

#ggsave("Figures/Figure_4A_density_acetate.jpeg", width = 16, height = 9, units = "in", dpi = 300)
```

#### B) Propionate

```{r}
#PROPIONATE
#Check if follows a normal distribution 
#Healthy donors 
shapiro.test(subset(propionate_data_bct, patient_type == "healthy_donor")$scfa_concentration) 
#NOT a normal distribution (p-value = 0.0003536)

#Dysbiotic patients
shapiro.test(subset(propionate_data_bct, patient_type == "cdiff_patient")$scfa_concentration) 
#NOT a normal distribution (p-value = 1.969e-07)

#DENSITY PLOT PROPIONATE
figure_4B_plot <- ggplot(propionate_data_bct, aes(x = scfa_concentration, fill = patient_type, label = labels_density)) +
  geom_density(alpha = 0.5, color = "dimgray", linewidth = 0)  +
  geom_vline(xintercept = roc_parameters_propionate$threshold, color = "dimgray", linetype = "dashed", size = 1) +
  geom_textdensity(linewidth = 1, color = "gray30", size = 8, hjust = 0.4, spacing = 40) +
  scale_fill_manual(values = c("healthy_donor" = "khaki", "cdiff_patient" = "khaki3")) +
  labs(x = "Propionate levels normalised by bacterial count", y = "Density", fill = NULL, tag = "B)") +
  scale_x_continuous(limits = c(0, 250), breaks = c(0, 50, 100, 150, 200, 250)) +
  scale_y_continuous(limits = c(0, 0.02), breaks = c(0, 0.005, 0.010, 0.015, 0.02)) +
  theme_minimal() +
  theme(legend.position = "none",
        text = element_text(size = 25),
        axis.text = element_text(size = 25),  # Size of axis labels
        axis.title.y = element_text(vjust = +3, size = 30), #Size of axis titles 
        axis.title.x = element_text(vjust = -1, size = 30), 
        strip.text.x = element_blank(), 
        plot.margin = margin(0.2, 0.2, 0.2, 0.2, "in")) #Margins of the plot 

figure_4B_plot

#ggsave("Figures/Figure_4B_density_propionate.jpeg", width = 16, height = 9, units = "in", dpi = 300)
```

#### C) Butyrate

```{r}
#BUTYRATE
#Check if follows a normal distribution 
#Healthy donors 
shapiro.test(subset(butyrate_data_bct, patient_type == "healthy_donor")$scfa_concentration) 
#NOT a normal distribution (p-value = 8.109e-05)

#Dysbiotic patients
shapiro.test(subset(butyrate_data_bct, patient_type == "cdiff_patient")$scfa_concentration) 
#NOT a normal distribution (p-value = 5.201e-08)

#DENSITY PLOT BUTYRATE
figure_4C_plot <- ggplot(butyrate_data_bct, aes(x = scfa_concentration, fill = patient_type, label = labels_density)) + 
  geom_density(alpha = 0.4, color = "dimgray", linewidth = 0)  +
  geom_vline(xintercept = roc_parameters_butyrate$threshold, color = "dimgray", linetype = "dashed", size = 1) +
  geom_textdensity(linewidth = 1, color = "gray30", size = 8, hjust = 0.17, spacing = 40) +
  scale_fill_manual(values = c("healthy_donor" = "#1F9A8AFF", "cdiff_patient" = "#24868EFF")) +
  labs(x = "Butyrate levels normalised by bacterial count", y = "Density", fill = NULL, tag = "C)") +
  scale_x_continuous(limits = c(0, 250), breaks = c(0, 50, 100, 150, 200, 250)) +
  scale_y_continuous(limits = c(0, 0.04), breaks = c(0, 0.01, 0.02, 0.03, 0.04)) +
  theme_minimal() +
  theme(legend.position = "none",
        text = element_text(size = 25),
        axis.text = element_text(size = 25),  # Size of axis labels
        axis.title.y = element_text(vjust = +3, size = 30), #Size of axis titles 
        axis.title.x = element_text(vjust = -1, size = 30), 
        strip.text.x = element_blank(), 
        plot.margin = margin(0.2, 0.2, 0.2, 0.2, "in")) #Margins of the plot 

figure_4C_plot

#ggsave("Figures/Figure_4C_density_butyrate.jpeg", width = 16, height = 9, units = "in", dpi = 300)
```

#### D) ROC curve

```{r}
#ROC curve with ggplot (geom_roc)
figure_4D_plot <- ggplot(scfa_bct_normalised, aes(d = roc_outcome, m = scfa_concentration, color = scfa)) + 
  geom_roc(n.cuts = 0, labels = FALSE, linealpha = 0.7, size = 2) + 
  geom_abline(intercept=0.0, slope = 1, linetype ="dashed", color = "gray", linewidth = 1) +
  scale_color_manual(values = c("acetate" = "#440154FF", "propionate" = "khaki3", "butyrate" = "#24868EFF"),
                    name = NULL, 
                    breaks = c("acetate", "propionate", "butyrate"), 
                    labels = c("Acetate - AUC-ROC = 0.696", "Propionate - AUC-ROC = 0.755", "Butyrate - AUC-ROC = 0.860")) + 
  scale_x_continuous("False Positive Rate (1 - Specificity)", breaks = seq(0, 1, by = .2)) + 
  scale_y_continuous("True Positive Rate (Sensitivity)", breaks = seq(0, 1, by = .2)) + 
  labs(tag = "D)") +
  theme_minimal() +
  theme(legend.position = c(0.745, 0.2), 
        legend.box.background = element_rect(color = "gray", fill = "white"),
        legend.key.width = unit(0.5, "in"), #Adjust the width of the legend box
        legend.key.height = unit(0.5, "in"),
        text = element_text(size = 25),
        axis.text = element_text(size = 25),  # Size of axis labels
        axis.title.y = element_text(vjust = +3, size = 30), #Size of axis titles 
        axis.title.x = element_text(vjust = -1, size = 30), 
        strip.text.x = element_blank(), 
        plot.margin = margin(0.2, 0.2, 0.2, 0.2, "in")) #Margins of the plot 

figure_4D_plot

#ggsave("Figures/Figure_4D_ROC_curve.jpeg", width = 16, height = 9, units = "in", dpi = 300)

#Calculate the AUC of each SCFA (they are the same as pROC)
auc_curve <- round(calc_auc(figure_4D_plot)["AUC"], 3)
auc_curve
```

## SUPPLEMENTARY MATERIALS

### Supplementary Figure S2. Combined ROC

Code for the plot of supplementary figure 2:

```{r}
#Combined dataframe with normalised SCFAs levels by the bacterial count, and combined according to gut proportions (60:20:20).
scfa_combined <- data.frame(acetate_data_bct$sample_number) 
scfa_combined <- as.data.frame(scfa_combined)
colnames(scfa_combined) <- c("patient_code")

scfa_combined <- mutate(scfa_combined, acetate_60 = rep(c(acetate_data_bct$scfa_concentration)*0.6, length.out = c(155))) %>% mutate(propionate_20 = rep(c(propionate_data_bct$scfa_concentration)*0.2, length.out = c(155))) %>% mutate(butyrate_20 = rep(c(butyrate_data_bct$scfa_concentration)*0.2, length.out = c(155))) %>% mutate(combined_scfa = acetate_60+propionate_20+butyrate_20) %>% mutate(roc_outcome = rep(c(1, 0), times = c(115, 40))) %>% mutate(scfa = "scfa_combination", length.out = c(155)) %>% dplyr::select (patient_code, combined_scfa, roc_outcome, scfa)

#ROC curve (pROC library)
#SCFAs (Data normalised by the bacterial count and considering SCFAs proportions)
#Assuming --> 1 = Healthy donors; 0 = Ill patients 
roc_obj_scfa <- roc(response = scfa_combined$roc_outcome, 
                       predictor = scfa_combined$combined_scfa)

#ROC parameters --> Acetate 
#pvalue
roc_area_scfa <- roc.area(as.numeric(as.vector(scfa_combined$roc_outcome)), scfa_combined$combined_scfa)
roc_area_scfa$p.value #p-value = 1.702708e-05

roc_parameters_scfa <- coords(roc_obj_scfa, "best", ret =c("threshold", "specificity", "sensitivity", "precision", "youden", "closest.topleft"), transpose = FALSE) %>% mutate(auc = roc_obj_scfa$auc, ci_lower_bound = ci(roc_obj_scfa, of = "auc", method = "delong")[1], ci_upper_bound = ci(roc_obj_scfa, of = "auc", method = "delong")[3]) %>% mutate(pvalue = roc_area_scfa$p.value)

#ROC curve with ggplot (geom_roc)
figure_S2_plot <- ggplot(scfa_combined, aes(d = roc_outcome, m = combined_scfa, colour = scfa)) + 
  geom_roc(n.cuts = 0, labels = FALSE, linealpha = 0.7, size = 2)  + 
  geom_abline(intercept=0.0, slope = 1, linetype ="dashed", color = "gray", linewidth = 0.4) +
  scale_color_manual(values = c("scfa_combination" = "green4"),
                    name = NULL, 
                    labels = c("AUC-ROC = 0.720")) +   
  scale_x_continuous("False Positive Rate (1 - Specificity)", breaks = seq(0, 1, by = .2)) + 
  scale_y_continuous("True Positive Rate (Sensitivity)", breaks = seq(0, 1, by = .2)) + 
  theme_minimal(base_size = 14) + 
  theme(legend.position = c(0.865, 0.14), 
        legend.box.background = element_rect(color = "gray", fill = "white"),
        legend.key.width = unit(0.5, "in"), #Adjust the width of the legend box
        legend.key.height = unit(0.5, "in"),
        text = element_text(size = 25),
        axis.text = element_text(size = 25),  # Size of axis labels
        axis.title.y = element_text(vjust = +3, size = 30), #Size of axis titles 
        axis.title.x = element_text(vjust = -1, size = 30), 
        strip.text.x = element_blank(), 
        plot.margin = margin(0.2, 0.2, 0.2, 0.2, "in")) #Margins of the plot 

figure_S2_plot

#Calculate the AUC of each SCFA (they are the same as pROC)
auc_curve_combined <- round(calc_auc(figure_S2_plot)["AUC"], 3)
auc_curve_combined

ggsave("Figures/Figure_S2_combined_ROC.jpeg", width = 16, height = 9, units = "in", dpi = 300)
```

### Supplementary Figure S3. PR curve

Code for the plot of supplementary figure 3:

We test the precision-recall curves. We can use the already existing predicted probabilities calculated in the ROC objects of the previous section of the analysis.

```{r}
#PR CURVES (PRROC library)
#Generate the PR curve using the same predictor in the ROC curves

#ACETATE
pr_acetate <- pr.curve(
  scores.class0 = acetate_data_bct$scfa_concentration[acetate_data_bct$roc_outcome == 1], 
  scores.class1 = acetate_data_bct$scfa_concentration[acetate_data_bct$roc_outcome == 0], 
  curve = TRUE)

#Extraction of the precision, recall, and thresholds
pr_data_acetate <- data.frame(Recall = pr_acetate$curve[, 1], Precision = pr_acetate$curve[, 2], Threshold = pr_acetate$curve[, 3]) %>% mutate(scfa = rep(c("acetate"), times = c(157)))

#AUC value for the PR curve
pr_auc_acetate <- pr_acetate$auc.integral #0.833

#PROPIONATE
pr_propionate <- pr.curve(
  scores.class0 = propionate_data_bct$scfa_concentration[propionate_data_bct$roc_outcome == 1],
  scores.class1 = propionate_data_bct$scfa_concentration[propionate_data_bct$roc_outcome == 0],
  curve = TRUE)

pr_data_propionate <- data.frame(Recall = pr_propionate$curve[, 1], Precision = pr_propionate$curve[, 2], Threshold = pr_propionate$curve[, 3]) %>% mutate(scfa = rep(c("propionate"), times = c(156)))

pr_auc_propionate <- pr_propionate$auc.integral #0.848

#BUTYRATE
pr_butyrate <- pr.curve(
  scores.class0 = butyrate_data_bct$scfa_concentration[butyrate_data_bct$roc_outcome == 1],
  scores.class1 = butyrate_data_bct$scfa_concentration[butyrate_data_bct$roc_outcome == 0],
  curve = TRUE)

pr_data_butyrate <- data.frame(Recall = pr_butyrate$curve[, 1], Precision = pr_butyrate$curve[, 2], Threshold = pr_butyrate$curve[, 3]) %>% mutate(scfa = rep(c("butyrate"), times = c(156)))

pr_auc_butyrate <- pr_butyrate$auc.integral #0.931

#Combine the data for the plot 
combined_pr_data <- rbind(pr_data_acetate, pr_data_propionate, pr_data_butyrate)

#PLOT OF THE PRECISION-RECALL CURVE
figure_S3_PR_curve <- ggplot(combined_pr_data, aes(x = Recall, y = Precision, color = scfa)) +
  geom_line(size = 1) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray", linewidth = 1) +  #Reference line
  scale_color_manual(values = c("acetate" = "#440154FF", "propionate" = "khaki3", "butyrate" = "#24868EFF"), 
                     name = NULL, 
                     breaks = c("acetate", "propionate", "butyrate"), 
                     labels = c("Acetate - AUC-PR = 0.833", "Propionate - AUC-PR = 0.848", "Butyrate - AUC-PR = 0.931")) + 
  scale_x_continuous("Recall", limits = c(0, 1), breaks = seq(0, 1, by = .2)) +
  scale_y_continuous("Precision", limits = c(0, 1),  breaks = seq(0, 1, by = .2)) + 
  theme_minimal() +
  theme(legend.position = c(0.84, 0.185), 
        legend.box.background = element_rect(color = "gray", fill = "white"),
        legend.key.width = unit(0.5, "in"), #Adjust the width of the legend box
        legend.key.height = unit(0.5, "in"),
        text = element_text(size = 25),
        axis.text = element_text(size = 25),  # Size of axis labels
        axis.title.y = element_text(vjust = +3, size = 30), #Size of axis titles 
        axis.title.x = element_text(vjust = -1, size = 30), 
        strip.text.x = element_blank(), 
        plot.margin = margin(0.2, 0.2, 0.2, 0.2, "in")) #Margins of the plot 

figure_S3_PR_curve

#ggsave("Figures/Figure_S3_PR_curve.jpeg", width = 16, height = 9, units = "in", dpi = 300)
```
